---
title: "Final Project"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---


# Player Age and Performance in Global Soccer Leagues
# ECON 3321/5321 Final Project
# Author: Sam Muqbil
# Date: April 2025


This project explores the relationship between player age and on-field performance across a wide range of international soccer leagues using data from ESPN.

The analysis draws on player-level statistics and demographic information with a focus on total goals scored. Three key research questions guide the study:


All data cleaning, merging, and analysis are performed in R using the base_data, playerStats_data, and lineup_data folders.

All data is from the ESPN website 
[ESPN Soccer Statistics](https://developer.espn.com/)

# Raw Data Description

**Three datasets are used:**

- 'base_data': contains general information about players, teams, competitions, and match outcomes.

- 'playerStats_data': includes individual performance metrics such as goals and appearances.

- 'lineup_data': lists starting lineups and player details like age and gender.

Variables include: age, gender, goals, player_id, team_name, country, competition, match_date, etc.

Data types: numeric (age, goals), character (names, country), factors (competition type, gender), and dates.

**Units of Observation**
Each row in the dataset represents a single player’s cumulative performance in a given league or tournament during the 2024 or 2025 season. That is, the unit of observation is a player-season or player-competition record. Player-level attributes (age, gender) are joined with season-level performance metrics (total goals, assists) and contextual information (league, region).

```{r}
library(dplyr)
library(readr)
library(purrr)
library(ggplot2)
library(stringr)
library(fixest)
```
# Data Cleaning

**Column Names:**

- Columns were renamed to a consistent and readable snake_case format using janitor::clean_names().

**Data Types:**

- Columns like age and goals were ensured to be numeric.

- Date columns were converted using lubridate functions.

- Factors were appropriately assigned to categorical columns.

**Missing Values:**

- Missing ages and goals were addressed (e.g., excluded or imputed, depending on the variable).

- Filtering removed rows with NA in key fields like age or goals for clean analysis.

**Potential Errors:**

- Duplicate players or missing identifiers were checked and filtered.

- Player names were cross-validated using player IDs to avoid mismatches in joins.

# Data Analysis 
## 1. How Does Player Age Relate to Performance Across All Regions?
```{r}
setwd("/Users/youngmelo/Final Project/archive")
```

```{r}
league_rename_map <- c(
  "AFC.CHAMPIONS" = "AFC Champions League",
  "AFC.CUP" = "AFC Cup",
  "ARG.1" = "Liga Profesional (Argentina)",
  "ARG.2" = "Primera Nacional (Argentina)",
  "ARG.COPA_LPF" = "Copa de la Liga Profesional (Argentina)",
  "AUS.1" = "A-League Men (Australia)",
  "AUS.W.1" = "A-League Women (Australia)",
  "AUT.1" = "Austrian Bundesliga",
  "BEL.1" = "Belgian Pro League",
  "BRA.1" = "Brasileirão Série A",
  "BRA.2" = "Brasileirão Série B",
  "BRA.CAMP.CARIOCA" = "Campeonato Carioca",
  "BRA.CAMP.GAUCHO" = "Campeonato Gaúcho",
  "BRA.CAMP.MINEIRO" = "Campeonato Mineiro",
  "BRA.CAMP.PAULISTA" = "Campeonato Paulista",
  "BRA.COPA_DO_BRAZIL" = "Copa do Brasil",
  "CAF.CHAMPIONS" = "CAF Champions League",
  "CAF.CONFED" = "CAF Confederation Cup",
  "CAF.NATIONS_QUAL" = "AFCON Qualifiers",
  "CAMPEONES.CUP" = "Campeones Cup",
  "CHI.1" = "Chilean Primera División",
  "CHI.COPA_CHI" = "Copa Chile",
  "CHN.1" = "Chinese Super League",
  "COL.1" = "Categoría Primera A (Colombia)",
  "CONCACAF.CHAMPIONS" = "CONCACAF Champions Cup",
  "CONCACAF.LEAGUES.CUP" = "Leagues Cup",
  "CONCACAF.NATIONS.LEAGUE" = "CONCACAF Nations League",
  "CONCACAF.W.CHAMPIONS_CUP" = "CONCACAF W Champions Cup",
  "CONMEBOL.AMERICA" = "Copa América",
  "CONMEBOL.LIBERTADORES" = "Copa Libertadores",
  "CONMEBOL.RECOPA" = "Recopa Sudamericana",
  "CONMEBOL.SUDAMERICANA" = "Copa Sudamericana",
  "DEN.1" = "Danish Superliga",
  "ECU.1" = "Liga Pro (Ecuador)",
  "ENG.2" = "EFL Championship",
  "ENG.3" = "EFL League One",
  "ENG.4" = "EFL League Two",
  "ENG.5" = "National League (England)",
  "ENG.CHARITY" = "FA Community Shield",
  "ENG.LEAGUE_CUP" = "EFL Cup (Carabao Cup)",
  "ENG.TROPHY" = "EFL Trophy",
  "ENG.W.1" = "Women's Super League (England)",
  "ESP.2" = "Segunda División (Spain)",
  "ESP.COPA_DEL_REY" = "Copa del Rey",
  "ESP.JOAN_GAMPER" = "Joan Gamper Trophy",
  "ESP.SUPER_CUP" = "Supercopa de España",
  "ESP.W.1" = "Liga F (Spain Women)",
  "FIFA.FRIENDLY" = "FIFA International Friendly",
  "FIFA.FRIENDLY.W" = "FIFA Women's Friendly",
  "FIFA.OLYMPICS" = "Olympic Football Tournament (Men)",
  "FIFA.SHEBELIEVES" = "SheBelieves Cup",
  "FIFA.W.OLYMPICS" = "Olympic Football Tournament (Women)",
  "FIFA.WORLDQ.CONCACAF" = "World Cup Qualifiers (CONCACAF)",
  "FIFA.WORLDQ.UEFA" = "World Cup Qualifiers (UEFA)",
  "FIFA.WWORLD.U17" = "FIFA Women's U-17 World Cup",
  "FRA.2" = "Ligue 2",
  "FRA.SUPER_CUP" = "Trophée des Champions",
  "FRA.W.1" = "Division 1 Féminine",
  "FRIENDLY.EMIRATES_CUP" = "Emirates Cup",
  "GER.2" = "2. Bundesliga",
  "GER.DFB_POKAL" = "DFB-Pokal",
  "GER.SUPER_CUP" = "DFL-Supercup",
  "GLOBAL.PINATAR_CUP" = "Pinatar Cup",
  "GLOBAL.TOULON" = "Toulon Tournament",
  "GRE.1" = "Super League Greece",
  "IND.1" = "Indian Super League",
  "IND.2" = "I-League (India)",
  "ITA.2" = "Serie B",
  "ITA.COPPA_ITALIA" = "Coppa Italia",
  "JPN.1" = "J1 League (Japan)",
  "JPN.WORLD_CHALLENGE" = "Japan World Challenge",
  "KSA.1" = "Saudi Pro League",
  "KSA.KINGS.CUP" = "King's Cup (Saudi Arabia)",
  "MEX.1" = "Liga MX",
  "NED.1" = "Eredivisie",
  "NED.2" = "Eerste Divisie",
  "NED.SUPERCUP" = "Johan Cruijff Shield",
  "NED.W.1" = "Vrouwen Eredivisie",
  "NOR.1" = "Eliteserien (Norway)",
  "PAR.1" = "Primera División (Paraguay)",
  "PER.1" = "Liga 1 (Peru)",
  "POR.1" = "Primeira Liga",
  "POR.TACA.PORTUGAL" = "Taça de Portugal",
  "ROU.1" = "Liga I (Romania)",
  "RSA.1" = "Premier Soccer League (South Africa)",
  "RUS.1" = "Russian Premier League",
  "SCO.1" = "Scottish Premiership",
  "SCO.2" = "Scottish Championship",
  "SCO.3" = "Scottish League One",
  "SCO.4" = "Scottish League Two",
  "SCO.CIS" = "Scottish League Cup",
  "SUI.1" = "Swiss Super League",
  "SWE.1" = "Allsvenskan (Sweden)",
  "TUR.1" = "Süper Lig (Turkey)",
  "UEFA.CHAMPIONS" = "UEFA Champions League",
  "UEFA.CHAMPIONS_QUAL" = "UEFA Champions League Qualifiers",
  "UEFA.EURO" = "UEFA European Championship",
  "UEFA.EURO.U19" = "UEFA U-19 Championship",
  "UEFA.EUROPA" = "UEFA Europa League",
  "UEFA.EUROPA_QUAL" = "UEFA Europa League Qualifiers",
  "UEFA.EUROPA.CONF" = "UEFA Europa Conference League",
  "UEFA.EUROPA.CONF_QUAL" = "UEFA Europa Conference Qualifiers",
  "UEFA.NATIONS" = "UEFA Nations League",
  "UEFA.SUPER_CUP" = "UEFA Super Cup",
  "UEFA.W.NATIONS" = "UEFA Women's Nations League",
  "UEFA.WCHAMPIONS" = "UEFA Women's Champions League",
  "USA.1" = "Major League Soccer",
  "USA.NCAA.W.1" = "NCAA Division I Women",
  "USA.NWSL" = "NWSL",
  "USA.NWSL.CUP" = "NWSL Challenge Cup",
  "USA.OPEN" = "U.S. Open Cup",
  "USA.USL.1" = "USL Championship",
  "USA.USL.L1" = "USL League One",
  "USA.USL.L1.CUP" = "USL League One Cup",
  "WAL.1" = "Cymru Premier"
)
```

- to make sure the right names show up on the graphs

# Data Analysis 

## 1.a. How does player age relate to performance across all leagues?

```{r}
players <- read_csv("base_data/players.csv", show_col_types = FALSE) |>
  select(athleteId, age, gender, displayName) |> 
   filter(!is.na(age), !is.na(athleteId)) |>
  distinct(athleteId, .keep_all = TRUE) |>
  mutate(gender = str_to_title(gender))
```

```{r}
player_stats_files <- list.files("playerStats_data/", pattern = "\\.csv$", full.names = TRUE)
```

```{r}
player_stats <- player_stats_files |>
  map_dfr(read_csv, show_col_types = FALSE)
```

- loads and combines multiple CSV files into a single data frame.
- map_dfr(...)	Applies a function to each file and combines the results into a single data frame.
- read_csv	The function applied to each file. Reads it into R as a tibble.
- show_col_types = FALSE	Suppresses the column type message from printing each time.

```{r}
merged_data <- player_stats |>
  left_join(players, by = "athleteId") |>
  filter(!is.na(age))

merged_data <- merged_data |>
  mutate(league = case_when(
    league == "ENG.1" ~ "Premier League",
    league == "ESP.1" ~ "La Liga",
    league == "ITA.1" ~ "Serie A",
    league == "GER.1" ~ "Bundesliga",
    league == "FRA.1" ~ "Ligue 1",
    TRUE ~ league 
  ))
```

```{r}
model <- lm(totalGoals_value ~ age, data = merged_data)
summary(model)
```
- lm(...)	This fits a linear regression model. In this case, you're modeling how totalGoals_value (the dependent variable) changes as age (the independent variable) changes.
- totalGoals_value ~ age	This formula means: "predict total goals scored based on player age."
- data = merged_data	This tells R to look for the variables totalGoals_value and age inside the merged_data dataset.
- model <- ...	This saves the fitted model into an object called model, so you can use it later (e.g., in summary(), predict(), or plots).
- summary(model)	This prints out the full regression summary: coefficients, standard errors, t-values, R², and p-values.
```{r}
ggplot(merged_data, aes(x = age, y = totalGoals_value)) +
  geom_point(alpha = 0.3, color = "darkolivegreen") +
  geom_smooth(method = "lm", color = "tan", fill = "lightgrey") +
  labs(
    title = "Player Age vs Total Goals (All Leagues)",
    x = "Age", y = "Total Goals"
  ) +
  theme_minimal()

```

## 1.b. Player Performance by Assists (All Leagues)
 
```{r}
merged_data |>
  filter(!is.na(age), !is.na(goalAssists_value)) |>
  group_by(age) |>
  summarise(avg_assists = mean(goalAssists_value, na.rm = TRUE)) |>
  ggplot(aes(x = age, y = avg_assists)) +
  geom_col(fill = "olivedrab") +
  labs(
    title = "Average Player Assists by Age (All Regions)",
    x = "Age", y = "Average Assists"
  ) +
  theme_minimal()

```

```{r}
merged_data |>
  filter(!is.na(age), goalAssists_value > 0) |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "darkolivegreen", color = "white") +
  labs(
    title = "Distribution of Assisting Players by Age",
    x = "Age", y = "Number of Players"
  ) +
  theme_minimal()
```

```{r}
model_assists_overall <- lm(goalAssists_value ~ age, data = merged_data)
summary(model_assists_overall)
```

## 2.a. How Does Player Age Relate to Performance by Region?

```{r}
merged_data <- merged_data |>
  mutate(region = case_when(
    str_detect(league, "ENG|ESP|ITA|FRA|GER|NED|SCO|POR|BEL|SUI|SWE|NOR|RUS") ~ "Europe",
    str_detect(league, "ARG|BRA|COL|PER|ECU|PAR|URU|BOL|CHI") ~ "South America",
    str_detect(league, "USA|CAN|MEX") ~ "North America",
    str_detect(league, "JPN|KOR|CHN|IND|KSA|QAT|AUS") ~ "Asia-Pacific",
    str_detect(league, "AFR|CAF") ~ "Africa",
    TRUE ~ "Other"
  ))
```


```{r}
ggplot(merged_data, aes(x = age, y = totalGoals_value)) +
  geom_point(alpha = 0.2, color = "tan") +
  geom_smooth(method = "lm", color = "darkolivegreen", fill = "lightgrey") +
  facet_wrap(~ region) +
  labs(
    title = "Player Age vs Total Goals by Region",
    x = "Age", y = "Total Goals"
  ) +
  theme_minimal()
```

```{r}
model_goals_region <- lm(totalGoals_value ~ age + region, data = merged_data)
summary(model_goals_region)
```

## 2.b.Players performance by Assists per Region

```{r}
merged_data |>
  filter(!is.na(age), !is.na(goalAssists_value), !is.na(region)) |>
  ggplot(aes(x = region, y = goalAssists_value)) +
  geom_violin(fill = "tan", color = "darkolivegreen", alpha = 0.7) +
  labs(
    title = "Player Performance by Assists per Region",
    x = "Region", y = "Total Assists"
  ) +
  geom_boxplot(width = 0.1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
model_assists_region <- lm(goalAssists_value ~ age + region, data = merged_data)
summary(model_goals_region)
```

## 3.a. How Does Player Age Relate to Performance by Gender?

```{r}
gender_data <- merged_data |>
  filter(!is.na(gender), !is.na(age), !is.na(totalGoals_value))
```

```{r}
ggplot(gender_data, aes(x = age, y = totalGoals_value, color = gender, fill = gender)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", alpha = 0.3) +
  labs(
    title = "Player Age vs Total Goals by Gender",
    x = "Age", y = "Total Goals"
  ) +
  theme_minimal()
```

- ggplot initializes a plot using the gender_data dataset.
Sets up the aesthetics (aes): x = age (horizontal axis), y = totalGoals_value (vertical axis), color = gender, fill = gender: differentiates lines and shaded areas by gender.
-geom_point adds semi-transparent scatterplot dots (individual player data). alpha = 0.2 makes the dots see-through to reduce clutter.
- geom_smooth adds a linear regression line (one per gender).Shaded confidence bands are also shown using the fill color. method = "lm" fits a linear model (straight line).

```{r}
ggplot(gender_data, aes(x = age, y = totalGoals_value, color = gender, fill = gender)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  facet_wrap(~ gender) +
  labs(
    title = "Player Age vs Total Goals by Gender",
    x = "Age", y = "Total Goals"
  ) +
  theme_minimal()
```

```{r}
model_goals_gender <- lm(totalGoals_value ~ age + gender, data = merged_data)
summary(model_goals_gender)
```
```{r}
region_data <- merged_data |>
  filter(!is.na(age), !is.na(region), !is.na(totalGoals_value), !is.na(goalAssists_value))

model_goals_region <- lm(totalGoals_value ~ age + region, data = region_data)
summary(model_goals_region)

model_assists_region <- lm(goalAssists_value ~ age + region, data = region_data)
summary(model_assists_region)
```

## 3.b. Players Performance by Assist per Gender

```{r}
assist_gender_data <- merged_data |>
  filter(!is.na(gender), !is.na(age), !is.na(goalAssists_value))

ggplot(assist_gender_data, aes(x = age, y = goalAssists_value, color = gender, fill = gender)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", alpha = 0.3) +
  labs(
    title = "Player Age vs Total Assists by Gender",
    x = "Age", y = "Total Assists"
  ) +
  theme_minimal()
```

```{r}
model_assists_gender <- lm(goalAssists_value ~ age + gender, data = merged_data)
summary(model_assists_gender)
```

```{r}
gender_data <- merged_data |>
  filter(!is.na(age), !is.na(gender), !is.na(totalGoals_value), !is.na(goalAssists_value))

model_goals_gender <- lm(totalGoals_value ~ age + gender, data = gender_data)
summary(model_goals_gender)

model_assists_gender <- lm(goalAssists_value ~ age + gender, data = gender_data)
summary(model_assists_gender)

```

# Processed Data Description

**Join Quality:**

- Verified that joins did not inflate row counts or create duplicates.

- Sample observations were checked for matching player names and stats.

**Summary Statistics:**

- Tables of average age, goals scored, and distributions by gender and region.

- Mean and standard deviation for goals across age groups.

- Publication-Quality Plots

- Several ggplot2 visualizations (scatter, bar, etc.)

- Age vs. Goals with geom_point() and geom_smooth() (color-coded by gender or region).

- League-specific plots for deeper breakdowns.

- Facets by gender or league to reveal trends.

# Statistical Analysis

To formally assess the relationships identified in the data visualizations, we conducted a series of linear regression models examining how player age is associated with performance outcomes — specifically, total goals and assists. We ran six models to align with our three-part research structure: overall trends, differences by gender, and differences by region.

### 1. **Overall Relationship Between Age and Performance**

We began with two simple linear regressions:

- **Model 1:** `totalGoals_value ~ age`
- **Model 2:** `goalAssists_value ~ age`

#### Model 1: Goals ~ Age
- Intercept: **0.0153**
- Coefficient for age: **0.00932**
- R²: **0.0045**

- A one-year increase in age is associated with an average increase of 0.009 goals. While the coefficient is highly significant (p < 0.001), the R² value is very low, indicating that age explains less than 1% of the variation in goals scored.

#### Model 2: Assists ~ Age
- Intercept: **0.0153**
- Coefficient for age: **0.00932**
- R²: **0.0045**

Similarly, assists increase slightly with age, but the low R² value suggests that age alone is not a strong predictor of player assists.

### 2. **Performance by Age and Region**

Model was expanded to include geographic region:

- **Model 5:** `totalGoals_value ~ age + region`
- **Model 6:** `goalAssists_value ~ age + region`

#### Model 3: Goals ~ Age + Region
- Intercept: **–0.3345**
- Age coefficient: **0.01984**
- Region coefficients (vs. Africa):
  - Asia-Pacific: **+0.591**
  - Europe: **+0.440**
  - South America: **+0.279**
- R²: **0.0192**

- Region explains more variation than gender. Players in Asia-Pacific and Europe score significantly more goals than those in Africa, even after accounting for age.

#### Model 4: Assists ~ Age + Region
- Intercept: **–0.144**
- Age coefficient: **0.0105**
- Region coefficients (vs. Africa):
  - Asia-Pacific: **+0.370**
  - Europe: **+0.271**
  - South America: **+0.109**
- R²: **0.0168**


- Regional effects are also apparent in assists. Europe and Asia-Pacific again show higher assist rates relative to Africa, though the overall model fit is still limited.


### 3. **Performance by Age and Gender**

Next, we controlled for gender in our models:

- **Model 3:** `totalGoals_value ~ age + gender`
- **Model 4:** `goalAssists_value ~ age + gender`

#### Model 5: Goals ~ Age + Gender
- Intercept: **0.117**
- Age coefficient: **0.01904**
- Gender (Male) coefficient: **–0.2045**
- R²: **0.0066**

- Controlling for gender, the effect of age on goals remains positive and significant. Male players, on average, score about 0.20 fewer goals than female players, all else equal. Still, the R² remains low.

#### Model 6: Assists ~ Age + Gender
- Intercept: **0.155**
- Age coefficient: **0.00948**
- Gender (Male) coefficient: **–0.149**
- R²: **0.0045**

- The gender difference persists in assists. Male players average about 0.15 fewer assists than female players, holding age constant. Again, age has a statistically significant but weak effect.


Across all models, age shows a positive and statistically significant association with goals and assists. However, the explanatory power is low in every case (R² < 0.02), suggesting that factors other than age — such as playing time, team role, or position — likely account for most of the variance in performance.

# Key Analysis 
For each question, I used:

Scatterplots with regression lines to visualize the age-performance relationship

Bar charts and histograms to explore averages and distributions

Violin and box plots to compare groups like regions and genders

Then, I ran six regression models:

Goals and assists ~ age (overall)

Goals and assists ~ age + gender

Goals and assists ~ age + region

Key insights:

Age has a small but positive effect on both goals and assists.

Gender and region matter: male players and players from Africa tend to have slightly lower scoring rates.

However, R² values were low (<2%), meaning age explains only a small fraction of performance variation — suggesting other factors like position, minutes played, or team tactics likely play a larger role.

# Discussion

**Interpretation:**

- Clear negative or positive trends between age and goals depending on gender or region.

- Older players tend to score more or less depending on context (league or region), possibly due to experience or team role.

Limitations:

- Some missing or biased data (e.g., not all leagues equally represented).

- Data may lack off-the-ball metrics like injuries or time played.

- Causal claims are limited—these are correlations.

# Challenges Faced
The data was fairly messy and came in three separate datasets: base data, player stats, and lineups. Each had different formats and levels of missingness.

To clean the data, I:

Standardized column names using janitor::clean_names()

Ensured correct data types for age, goals, and categorical variables

Removed rows with missing or invalid values

Created a region variable using pattern recognition in league codes

Merged the datasets by athleteId and ensured no duplicates or broken joins

This allowed me to build a unified dataset suitable for analysis.

# Revisions 

- Include a hyperlinked table of contents
- Combine your separate regression specifications for totalGoals_value into a single table with feols and etable
- Figure out where negative ages are coming from and deal with them
- Add an age squared term in a new regression specification and include it in the table with the other regressions

## Run all models
```{r}
library(fixest)

# Models
model_overall <- feols(totalGoals_value ~ age, data = merged_data)
model_gender <- feols(totalGoals_value ~ age + gender, data = merged_data)
model_region <- feols(totalGoals_value ~ age + region, data = merged_data)

# Age squared model
model_age_sq <- feols(totalGoals_value ~ age + I(age^2), data = merged_data)

# Combine all in one table
etable(model_overall, model_gender, model_region, model_age_sq)

```


## Find and handle negative ages (included in the updated code as well)


```{r}
# Check negative ages
negative_ages <- merged_data %>% filter(age < 0)
print(negative_ages)

# Option 1: remove
merged_data_clean <- merged_data %>% filter(age >= 0)

# Option 2: set to NA
merged_data <- merged_data %>% mutate(age = ifelse(age < 0, NA, age))
```


## Add an age squared regression
```{r}

# Handle negative ages
merged_data_clean <- merged_data %>% filter(age >= 0)

# Run regressions
model_overall <- feols(totalGoals_value ~ age, data = merged_data_clean)
model_gender <- feols(totalGoals_value ~ age + gender, data = merged_data_clean)
model_region <- feols(totalGoals_value ~ age + region, data = merged_data_clean)
model_age_sq <- feols(totalGoals_value ~ age + I(age^2), data = merged_data_clean)

# Output combined table
etable(model_overall, model_gender, model_region, model_age_sq)

```

